---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rtranscribe

<!-- badges: start -->
[![R-CMD-check](https://github.com/JBGruber/rtranscribe/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JBGruber/rtranscribe/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

R bindings to [transcribe.cpp](https://github.com/handy-computer/transcribe.cpp), a C++ speech-recognition library that runs 16 model families (Whisper, Parakeet, Moonshine, Voxtral, Canary and more) locally on the CPU.

Everything runs on your machine: no API keys, no per-minute billing, and no audio leaves the computer.
The C++ sources are bundled and compiled at install time, so there is no separate library to install.

## Installation

You need **cmake** and a C++17 compiler.
The R sources build the bundled library, which takes a few minutes the first time.

``` r
# install.packages("remotes")
remotes::install_github("JBGruber/rtranscribe")
```

## Quick start

```{r}
library(rtranscribe)

# Fetch a small model (44 MB) into the package cache
model <- transcribe_download_model("whisper-tiny")

# Transcribe (uses a small audio file included in the package)
jfk_file <- system.file("extdata", "jfk.wav", package = "rtranscribe")
res <- transcribe(jfk_file, model)

res$text
res
```

Results come back as tibbles:

```{r}
res$segments
res$words # word-level rows, when the model produces them
res$tokens # token ids with confidences
res$speakers # who spoke when, after diarization
```

Any audio format `ffmpeg` reads works (wav, mp3, m4a, flac, ogg, and the audio track of video files) via the [av](https://cran.r-project.org/package=av) package.
You can also pass a numeric vector of 16 kHz mono PCM directly.

## Choosing a model

```{r}
transcribe_models()
```

That is a curated shortlist.
`refresh = TRUE` fetches the full catalogue (~70 models) from the Hugging Face API and appends everything else at the bottom:

```{r}
transcribe_models(refresh = TRUE)
```

The appended rows carry only a name: the listing API does not report size, WER or the transcribe.cpp architecture, and the repository tags are not a safe substitute (`moonshine-streaming-*` is tagged `moonshine`, but its family is `moonshine_streaming`).
Read the authoritative architecture with `transcribe_model_info()` after downloading.
Once refreshed, any listed name can be passed to `transcribe_download_model()`.

This is a convenience, not a limit: any GGUF converted for transcribe.cpp works.
Browse the catalogue at [huggingface.co/handy-computer](https://huggingface.co/handy-computer) and pass a path or URL directly.

``` r
m <- transcribe_load_model("~/models/parakeet-tdt-0.6b-v3-Q8_0.gguf")
transcribe_capabilities(m)
transcribe_supports(m, "diarization")
```

## Reusing a model

Loading a model is the expensive part.
Load once, then reuse the session:

```{r}
m <- transcribe_load_model(model)
s <- transcribe_session(m, n_threads = 8)

# assuming that the folder audio contains audio files
for (f in list.files("audio", full.names = TRUE)) {
  res <- transcribe_run(s, f)
  cat(basename(f), ":", res$text, "\n")
}
```

For many short clips, `transcribe_run_batch()` processes them in one dispatch:

```{r}
files <- list.files("inst/extdata", pattern = "\\.wav$", full.names = TRUE)
results <- transcribe_run_batch(s, files)
results
```

## Timestamps, translation and speakers

Timestamps, translation and diarization are three independent capabilities and no single family has all three, so this section uses three sessions.
Whisper tops out at segment timestamps; Parakeet resolves timings per token but has no translate head; speaker attribution needs a model trained for it.

```{r}
# Word-level timings: Parakeet times every token, so "word" is available
pm <- transcribe_download_model("parakeet-tdt-0.6b-v3") |>
  transcribe_load_model()
ps <- transcribe_session(pm, n_threads = 8)
res <- transcribe_run(ps, jfk_file, timestamps = "word")
res$words

# Translate into English. Note the model: the turbo distillation dropped the
# translate task, so this needs a plain multilingual whisper.
tm <- transcribe_download_model("whisper-base") |>
  transcribe_load_model()
german_file <- system.file("extdata", "german.wav", package = "rtranscribe")
transcribe_run(
  transcribe_session(tm),
  german_file,
  task = "translate",
  target_language = "en"
)
```

```r
# Speaker attribution, on a model that advertises it
dm <- transcribe_download_model("moss-transcribe-diarize") |>
  transcribe_load_model()
res <- transcribe_run(transcribe_session(dm), "meeting.wav", diarize = TRUE)
res$speakers
res$segments$speaker_id
```

Asking a model for something it cannot do is an error rather than a silent downgrade -- `unsupported timestamp granularity (status 12)` for timestamps, `unsupported task (status 11)` for translation -- so check the model first:

```{r}
transcribe_capabilities(pm)$max_timestamp_kind # "token"
transcribe_capabilities(m)$max_timestamp_kind # "segment"
transcribe_capabilities(m)$supports_translate # FALSE, on turbo
transcribe_supports(pm, "diarization")
```

Both limits are properties of the model, not of its size.
Every Whisper variant caps at `"segment"`, `whisper-large-v3-turbo` included; Parakeet, GigaAM and MedASR reach `"token"`, `granite-speech-*-plus` reaches `"word"`, and the remaining families report `"none"`.
`timestamps = "auto"` (the default) never errors: it resolves to whatever the model can produce, and `res$timestamp_kind` reports what you actually got.

## Model-specific options

Some families expose knobs that do not generalise.
They are passed with `family =` and validated against the loaded model:

```{r}
video_f <- curl::curl_download(
  "https://www.gesis.org/fileadmin/Home/GESIS-Imagefilm_en.webm",
  "GESIS-Imagefilm_en.webm"
)
# Bias Whisper's decoding toward domain vocabulary
transcribe_run(
  s,
  video_f,
  family = whisper_options(
    initial_prompt = "GESIS – Leibniz Institute for the Social Sciences "
  )
)

transcribe_accepts_options(m, whisper_options()) # TRUE for whisper models
```

Also available: `parakeet_stream_options()`, `parakeet_buffered_stream_options()`, `moonshine_streaming_options()` and `voxtral_realtime_options()`.

## GPU backends

The default build is CPU-only, because it has to work on any machine with a compiler.
A GPU backend is compiled in on request, at install time:

```sh
TRANSCRIBE_R_VULKAN=1 R CMD INSTALL .   # AMD, Intel and NVIDIA
TRANSCRIBE_R_CUDA=1   R CMD INSTALL .   # NVIDIA only
TRANSCRIBE_R_METAL=1  R CMD INSTALL .   # Apple Silicon
```

The same variables work with `pak` and `remotes`, which build from source when installing from GitHub.
Set them in `~/.Renviron` (`usethis::edit_r_environ(scope = "project")`) rather than with `Sys.setenv()`: pak builds in a background process that inherits the environment when it starts, so a variable set later in the session may not reach the build.

```r
# ~/.Renviron:  TRANSCRIBE_R_VULKAN=1
pak::pak("JBGruber/rtranscribe")
```

Each backend needs its SDK at build time — Vulkan needs the loader headers and the `glslc` shader compiler, CUDA needs the toolkit — and `configure` checks for them up front, with a per-distribution install hint if something is missing.
More than one can be enabled in the same build.

Then use it:

```r
transcribe_devices()                     # every device the runtime can see
transcribe_backend_available("vulkan")   # TRUE only if a Vulkan device is there
m <- transcribe_load_model(path, backend = "vulkan")   # or "auto"
```

`transcribe_backend_available()` is a device probe, not a build probe: it needs both a build that compiled the backend in *and* a working driver, so it stays `FALSE` on a Vulkan build running on a machine with no Vulkan device.

`backend = "auto"` takes the first GPU device that initialises (discrete before integrated) and falls back to the CPU if none does, so a GPU build still runs on a machine with no usable GPU.
Naming a backend explicitly is an assertion: `backend = "vulkan"` errors rather than falling back.

Two caveats.
Prebuilt r-universe binaries are CPU-only, so a GPU build always means a source install.
And of the three, only Vulkan has been tested here — CUDA and Metal are wired up but unverified on hardware.
